---
title: "__Key Solution for Assignment1: Fall-2024__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1: (80 points) 
Use the `tq_get()` function in the package __tidyquant__ to get the stock prices (from Jan 01, 2018 to December 31, 2022) of the two crypto currencies: Bitcoin ("BTC-USD") and Ethereum ("ETH-USD"). Name your data as __df__ and use it to solve the following questions:

## 1.1) (10 points) 
Use the `mutate()` function from the __dplyr__ package along with the `ymd()` function from the __lubridate__ package to convert the `date` varaible to the format "year-month-day". 

### Solution: 

Step1: Load the packages that you need to solve this assignment.

```
lapply(c("tidyquant","lubridate","tidyverse","ggpubr", "dygraphs", "skimr","plotly"), 
       require, character.only = TRUE)
```
```{r,fig.width = 14,fig.height = 6,fig.align="center",warning=FALSE, message=FALSE, results= "hide", echo = FALSE}
lapply(c("tidyquant","lubridate","tidyverse","ggpubr", "dygraphs", "skimr","plotly"), 
       require, character.only = TRUE)
```

Step2: Use the function `tq_get()` to get the quantitative data in tibble format and name your data as __df__.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
df <- tq_get(c("BTC-USD","ETH-USD"),
             from = "2018-01-01", to = "2022-12-31",
             get = "stock.prices")
```

Step3: Use the `mutate()` function from the __dplyr__ package along with the `ymd()` function from the __lubridate__ package to parse dates with year, month, and day components

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
df <- df %>%  
  mutate(date = ymd(date))
```

## 1.2) (5 points) 
Use the `mutate()` function from the __dplyr__ package to create a new variable: $r_t = \log x_t -\log x_{t-1}$, where $r_t$ and $x_t$ are the log returns and closing price series, respectively, at time $t$. Name the new variables as _ret_. __Hint:__ To calculate $r_t = \log x_t -\log x_{t-1}$ for each currency, use the `case_when(symbol == )` function to differentiate between currencies. After computing the returns, apply `drop_na()` function to remove any `NA` values, and then use the `bind_rows()` function to combine the rows from both currencies.

### Solution: 
Create the new variable and then remove "NA" values. 

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
df1 <- df %>% filter(symbol == "BTC-USD") %>%
      mutate(ret = log(close) - log(dplyr::lag(close)))%>%
  drop_na(ret)

df2 <- df %>% filter(symbol == "ETH-USD") %>%
  mutate(ret = log(close) - log(dplyr::lag(close)))%>%
  drop_na(ret)

df <- bind_rows(df1, df2)
```

## 1.3) (10 points) 
Use the `group_by()` and `summarise()` functions from the __dplyr__ package to calculate the mean and quartiles (1st, 2nd, and 3rd) of the log returns for each currency. Provide a comparison of the mean values and the 1st and 3rd quartiles between the two currencies.

### Solution: 
First, use the `group_by()` function in the package __dplyr__ to group the data by the variable `symbol`, which denotes the name of the currency, and then use the `summarise()` function  to find the means and the quartiles (1st, 2nd, and 3rd) of the log returns.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
# Calculate the mean
df %>% group_by(symbol) %>% 
  summarise(mean = mean(ret))
# # Calculate the 1st, 2nd, and 3rd quartiles
df %>% group_by(symbol) %>% 
  summarise(q25 = quantile(ret, 0.25),
            q50 = quantile(ret, 0.5),
            q75 = quantile(ret, 0.75))
```

Positive log returns indicate that currencies closing price at the end of the period is higher than at the beginning. 

For both currencies, approximately 25\% of the days show a closing price that is lower at the end of the period compared to the beginning. Meanwhile, about 75\% of the days show a closing price that is higher at the end of the period than at the beginning.

## 1.4) (10 points) 
Use the `mutate()` function from the __dplyr__ package along with the `case_when()` from the same package and add a new categorical variable called `covid`. This variable should have 3 categories `"Pre COVID-19"` for dates before December 1, 2019, `"Covid-19"` for dates between December 1, 2019, and February 10, 2022, and `"Post COVID-19"` for dates after February 10, 2022. 

### Solution:
You can use the following code to create this categorical variable. 
Note you might need to use the syntax `across()` to control the order in which the covid panels are ordered as you need this in the plots later.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
df <- df %>% 
  mutate(covid = case_when(
    date < as.Date("2019-12-01") ~ "Pre COVID-19 ",
    date >= as.Date("2019-12-01") & date < as.Date("2022-02-10") ~ "Covid-19",
    date >= as.Date("2022-02-10") ~ "Post COVID-19 ",
    TRUE ~ NA_character_))
df <- df %>% 
  mutate(across(covid, factor, levels=c("Pre COVID-19 ","Covid-19","Post COVID-19 "))) 
``` 


## 1.5) (10 points)
Use the __skimr__ package to determine the minimum and maximum closing stock prices for both currencies during the COVID-19 period. Compare the minimum and maximum closing prices of Bitcoin during the COVID-19 period with those from the pre-COVID-19 and post-COVID-19 periods. 

### Solution: 
First, use the `select()` function to choose the three variables: `close`, `covid`, and `symbol`. Next, use the `group_by()` function to group the data by the `covid` and `symbol` variables. Finaly, apply the `skim()` function from the package __skimr__ to obtain a summary of the grouped data.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
df %>% 
  select(close,covid,symbol) %>% 
  group_by(covid,symbol) %>% 
  skim() 
```

From the skim summary, you notice that the minimum and maximum closing prices of Bitcoin during the pre covid-19 time were `$3236.7617` and `$17527.000`, respectively. 
During the covid-19 era, these values were increased to `$84.3083` and `$1396.420` and keep increased on the post covid-19 period. 

## 1.6) (10 points) 
Use the `facet_wrap()` function from the __ggplot2__ package to recreate the plots shown in Figure 1. Write tidy code to generate these plots and provide a commentary on your observations regarding the log returns.

### Solution:

To produce the plots in Figure 1, you need to use the `facet_wrap()` function. 
Note that you also need to control the order in which the `covid` panels are ordered in the plot, which you already did in solving part (1.4) of this assignment

```{r,out.width="100%",fig.align="center",warning=FALSE, message=FALSE}
df %>%
  ggplot(aes(covid, ret, color = symbol)) + 
    geom_boxplot(outlier.colour = "#3366FF", 
                 outlier.shape = 4) + 
    facet_wrap(~symbol, scales = "free") + 
  labs(x = "Covid time", y="Log returns") +
  theme_bw() 
```
We see that both currencies have number of outliers.

## 1.7) (10 points) 
Write a tidy code to reproduce Figure 2 below. 

### Solution:
Below the code that you may write to reproduce Figure 2. Note that you need to include the syntax `legend.position = "bottom"` to change the legend position.

```
df %>% 
  ggplot(aes(symbol, ret, color = symbol)) + 
    geom_boxplot(outlier.colour = "#3366FF", 
               outlier.shape = 4) + 
    facet_wrap(~covid, scales = "free") + 
  labs(x = "Currency", y="Log returns") + 
  theme(legend.position = "bottom") + 
  theme_bw() 
```

```{r,fig.width = 14,fig.height = 6,fig.align="center",warning=FALSE, message=FALSE, results= "hide", echo = FALSE}
df %>% 
  ggplot(aes(symbol, ret, color = symbol)) + 
    geom_boxplot(outlier.colour = "#3366FF", 
               outlier.shape = 4) + 
    facet_wrap(~covid, scales = "free") + 
  labs(x = "Currency", y="Log returns") + 
  theme(legend.position = "bottom") + 
  theme_bw() 
```

## 1.8) (10 points) 
Write a tidy code to reproduce the bar chart in Figure 3 below. Note that you need to create an interactive plot using the function `ggplotly()` from the package __plotly__. Comment what do you notice about the log returns for both currencies? 

```{r,fig.width = 8.7,fig.height = 5.1,fig.align="center",warning=FALSE, message=FALSE}
p <- df %>%
  ggplot(aes(symbol, ret, fill = symbol)) +
geom_bar(stat = "summary", fun.y = "mean") +
  labs(x = "Currency", y="Average Log returns") +
  facet_wrap(~covid) + 
  theme(legend.position="bottom") 
ggplotly(p) %>% #create an interactive plot
  layout(legend = list(orientation = "h"))
```

It is evident that during the pandemic, the returns for both currencies increased significantly, but they declined post-COVID-19. Overall, Ethereum's returns have generally been higher than those of Bitcoin during the covid era. 
## 1.9) (5 points) 
Use the function `dygraph()` in the package __dygraphs__ to produce a dynamic graphic of the returns of the two series. Add a shading effect to your graph background using a `color = "#d3d3d3"` for the covid-19 period.

### Solution:
First, you might need to use the functions `filter()` and `select()` to select the returns of both currencies and then you can create a new data frame with three columns, (say "date", "BTC", "ETH") as follows:

```{r,fig.width = 14,fig.height = 8,fig.align="center",warning=FALSE, message=FALSE}
BTC <- df %>% filter(symbol == "BTC-USD") %>% select(date,close)
ETH <- df %>% filter(symbol == "ETH-USD") %>% select(close)

df1 <- data.frame(BTC, ETH)
names(df1) <- c("date", "BTC", "ETH")
```

Next, you can use the following code to get the dynamic graphic as requested.
```{r,out.width="100%",fig.align="center",warning=FALSE, message=FALSE}
dygraph(df1, main = "Log returns") %>%
  dyShading(from = "2019-12-01", 
            to = "2022-02-10", 
            color = "#d3d3d3") %>%
  dyOptions(colors = c("black",
                       "red")
            )
```

# Q2: (20 points) 
Use the `set.seed(123)` and the function `tibble()` to generate a random date of dimension $2000\times 6$ with the following variables and do some summary statistics:

-    Gender: (nominal with two categories "male" and "female") $40\%$ are males and $60\%$ are females. Use the `sample()` function.
-    Age: (numeric) use gamma-distribution with shape 50. Use the `rgamma()` function.
-    Height: (numeric) use normal distribution with mean 160 and standard deviation 8. Use the `rnorm`()` function.
-    Meal: (integer denotes the number of meals per day) random integers from 0 to 6. Use the `sample()` function.
-    Gym: (integer denotes the number of hours spent at the gym per day) use uniform distribution that ranges in value from 0 to 10 and round your generated data to integer numbers. Use the `runif()` function.
-    Status: (nominal denotes the marital status) single, married, and others. Use the `sample()` function.
-    Weight: (numeric) use the following equation:

\begin{aligned}
\text{Weight }&= 0.06 \text{ age }+ 0.7 \text{ height }+ 0.2 \text{ meal }- 0.1 \text{ male }- 0.16 \text{ gym }- 0.12 \text{ single } \\
&- 0.05 \text{ married }+ \text{ random error from t-distribution with 5 degrees of freedom}
\end{aligned}

## 2.1) (10 points)
Write the code that generates the data as described above.

### Solution:

```{r,out.width="100%",fig.align="center",warning=FALSE, message=FALSE}
set.seed(123)
gender <- sample(c("male","female"), size=2000, replace=TRUE, prob=c(0.6,0.4))
age <- rgamma(2000, shape = 50)
height <- rnorm(2000, mean = 160, sd = 8)
meal <- sample(0:6,2000, replace=T) 
gym<- round(runif(2000, min = 0, max = 10),0)
status <- sample(c("single", "married", "others"), 2000, replace=T)

weight <- 0.06 * age + 0.7 * height + 0.2 * meal - 0.16 * gym -
  - 0.1 * as.numeric(status == "male") +
  0.12 * as.numeric(status == "single") - 
  0.05 * as.numeric(status == "married") + 
    rt(2000, df = 5) #random data from t-distribution with df = 5 

sim_df <- tibble(gender, age, height, meal, gym, status, weight)
```

## 2.2) (5 points)
Find the age of the youngest single male who spends 3 hours at gym per a day.

### Solution:
To find the youngest single male who spends 3 hours at gym per a day, you need (optional step) first to group the data using the variables `gender`, `status`, and `gym` using the syntax `group_by()`, then you need (mandatory step) to apply the function `filter()` to select the rows that match the `gender = male`, `status = single`, and `gym = 3`, and lastly, you need to apply the function `summarise(min(age))` to get the youngest age as 31.1 years.

```{r,out.width="100%",fig.align="center",warning=FALSE, message=FALSE}
sim_df %>% group_by(gender, status, gym) %>%
  filter(gender == "male", status == "single", gym == 3) %>%
  summarise(min(age))
```

## 2.3) (5 points)
Suppose you need to provide a promotion for married customers aged between 40 and 60 who never go to the gym. Write a code which creates a new binary variable (name it as `promotion`) with two categories `yes, no`. What is the weight average of the those who will be eligible to get the promotion?  

### Solution:
You first need to mutate a new variable `promotion` which has `yes, no` values and then you find the mean weight for only those who will get the promotion as 115 kg. 
```{r,out.width="100%",fig.align="center",warning=FALSE, message=FALSE}
sim_df %>% 
  mutate(promotion = 
           if_else(status == "married" & 
                   gym == 0 & 
                   age > 40 & 
                   age < 60, "yes", "no")
         ) %>% 
  filter(promotion == "yes") %>%
  summarise(mean(weight))
```
