---
title: "__Bagging and Random Forests for Regression Trees__"
author: "Esam Mahdi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this example, we apply bagging and random forests to the <span  style="color:red">Boston</span> dataset, sourced from the __ISLR2__ library. Data has 506 rows and 13 variables:

-    <span style="color:green">crim</span>: per capita crime rate by town.

-    <span style="color:green">zn</span>: proportion of residential land zoned for lots over 25,000 sq.ft.

-    <span style="color:green">indus</span>: proportion of non-retail business acres per town.

-    <span style="color:green">chas</span>: Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).

-    <span style="color:green">nox</span>: nitrogen oxides concentration (parts per 10 million).

-    <span style="color:green">rm</span>: average number of rooms per dwelling.

-    <span style="color:green">age</span>: proportion of owner-occupied units built prior to 1940.

-    <span style="color:green">dis</span>: weighted mean of distances to five Boston employment centres.

-    <span style="color:green">rad</span>: index of accessibility to radial highways.

-    <span style="color:green">tax</span>: full-value property-tax rate per $10,000.

-    <span style="color:green">ptratio</span>: pupil-teacher ratio by town.

-    <span style="color:green">lstat</span>: lower status of the population (percent).

-    <span style="color:green">medv</span>: (Response variable) median value of owner-occupied homes in $1000s.

# Fitting Regression Trees

## Step 1: Data partition 

Load the required libraries and create training/test sets.
Here, we split the observations into a training set (say 50%) and a test data (say 50%). In this example, we use the <span style="color:blue">createDataPartition()</span> function in the __caret__ library. For more information about this function, type `?createDataPartition`.

```{r, warning=FALSE, message=FALSE}
# Load the required packages
library(ISLR2) # Introduction to Statistical Learning, Second Edition
library(caret) # Classification And REgression Training
library(rpart.plot) # Plot 'rpart' Models
# Create 50% training set and 50% test set (hold-out-validation)
set.seed(1)
Index <- createDataPartition(Boston[,"medv"], p = 0.5, list = FALSE)
# Alternatively, we may use the sample() function as below:  
# Index <- sample(1:nrow(Boston), size = nrow(Boston)/2)
train.data <- Boston[Index,]  # 254 observations
test.data <- Boston[-Index,]  # 252 observations
```

## Step 2: Model building 

Fit the tree to the training data using the <span style="color:blue">train()</span> function in the __caret__ library. 
The package __caret__ call the <span style="color:blue">rpart()</span> function from the package __rpart__ (Recursive Partitioning and Regression Trees) and train the model through cross-validation.
For more information about these functions, type `?rpart` and `?train`.
The important arguments of the <span style="color:blue">train()</span> function are given below:

-    <span style="color:green">form</span>: a formula that links the response (target) variable to the independent (features) variables. 

-    <span style="color:green">data</span>: the data to be used for modeling. 

-    <span style="color:green">trControl</span>: specify the control parameters for the resampling method and other settings when training a machine learning model. It is important when performing cross-validation or bootstrapping to assess the model's performance. 
Here are some of the sub-arguments we can set within `trControl`:
     -  <span style="color:green">method</span>: specifies the resampling method to be used, such as `"cv"` for k-fold cross-validation, `"repeatedcv"` for repeated k-fold cross-validation, `"boot"` for bootstrap resampling, and more.
     -  <span style="color:green">number</span>: it defines the number of folds in k-fold cross-validation ($=k$).
     -  <span style="color:green">repeats</span>: for repeated k-fold cross-validation, it specifies the number of times the cross-validation process is repeated.
     -  <span style="color:green">p</span>: For bootstrap resampling, it sets the proportion of data to resample in each iteration.
    
-    <span style="color:green">method</span>: defines the algorithm specifying which classification or regression model to use. In the case of decision trees we use the default method as `"rpart2"` where the model complexity is determined using the one-standard error method. 

-    <span style="color:green">preProcess</span>: an optional argument is used to perform mean centering and standardize predictor variables by typing `preProcess = c("center", "scale")`. 

-    <span style="color:green">tuneLength</span>: specifies the number of values to consider during the hyperparameter tuning process. It typically applies to the `cp` parameter, which controls the complexity of the tree in regression tree models. 

-    <span style="color:green">tuneGrid</span>: define a grid of values for the `cp` parameter.

```{r}
ctrl <- trainControl(method = "cv", number = 5) # use 5-fold cross-validation
fit1 <- train(medv ~., 
             data = train.data,
             method ="rpart2", # rpart2 is an enhanced version of rpart to handle overfitting more effectively
             trControl = ctrl,
             preProcess = c("center", "scale"),
             #tuneGrid = expand.grid(cp = seq(0.01,0.5, by = 0.01)),
             tuneLength = 10 # consider 10 different values of the cp (maxdepth = 10)
             ) 
fit1
plot(fit1)
```

After examining the output and the fitted model plot, it becomes evident that the root mean squared errors (RMSE) exhibit minimal variation when the maximum tree depth exceeds 6. Consequently, we establish the maximum depth to be 6 and proceed to retrain the model using the <span style="color:blue">rpart()</span> function from the __rpart__ library as shown below:

```{r}
fit2 <- rpart(medv ~., 
              train.data, 
              method  = "anova", #default method for regression trees
              maxdepth = 6) 
fit2
```
Note that the <span style="color:green">method</span> argument is used to define the algorithm that we use to fit the model. It can be one of `"anova"`, `"poisson"`, `"class"`, `"exp"`. In the case of regression trees where the target variable is numeric, we use the default method as `"anova"`.

Notice that the output of the fitted model indicates that only four of the variables have been used in constructing the tree (`lstat`, `nox`, `rad`, and `rm`).
In the context of a regression tree, the <span style="color:blue">deviance</span> is simply the sum of squared errors (SSE) for the tree.

The output of the fitted model shows the steps of the trees splits (root, branch, leaf).
For example, we start with 254 observations at the root node and we split the data first on the <span style="color:blue">lstat</span> variable (first branch). 
That is, out of all other features, <span style="color:blue">lstat</span> is the most predictive variable that optimizes a reduction in SSE. 

We see that 163 observations for which <span style="color:blue">lstat</span> are more than or equal to 8.96 will go to the left hand side of the first branch (denoted by <span style="color:red">2)</span>) and 91 observations for which <span style="color:blue">lstat</span> less than 8.96 will go to the right hand side of the first branch (denoted by <span style="color:red">3)</span>). 
The <span style="color:blue">asterisks (*)</span> indicate the leaf nodes associated with prediction values.   
For example, out of the 91 observations where $\text{lstat} < 8.96$, we see that 12 observations are following the terminal node (leaf) $\text{rm} \ge 7.4235$ with a predicted median value of owner-occupied homes (on average) of $\$45.74167$. The SSE of these observations (deviance) is $228.2292$.

## Step 3: Model visualizing 

We can easily visualize the model output by plotting the tree using the <span style="color:blue">rpart.plot()</span> function in the __rpart.plot__ library using the following code.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
rpart.plot(fit2, type = 2, digits = 3, fallen.leaves = FALSE)
```

## Step 4: Cost complexity pruning   

Note that the <span style="color:blue">rpart()</span> function is automatically applying a range of `cost complexity values and tuning parameter` ($\alpha$) to prune the tree that has the lowest error. 
This function performs a 10-fold cross validation, comparing the error that is associated with each $\alpha$ on the hold-out validation data. 
This process helps identify the optimal $\alpha$ and, consequently, the most suitable subtree to prevent overfitting the data.
To gain insight into the internal process, we can utilize the <span style="color:blue">printcp()</span> function as demonstrated below. For instance, with no split, the `cross-validation error (xerror)` is 1.00716, whereas with 6 splits, this error decreases to 0.29698.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
# Cost complexity pruning
printcp(fit2)
# Extract minimum error associated with the optimal cost complexity value for each model
cp <- fit2$cptable
```


We may also use the <span style="color:blue">plotcp()</span> function to create a graph displaying the cross-validation errors (y-axis) in relation to the cost complexity (CP) values (x-axis).
This plot is useful to choose the optimal value for `cost complexity (CP)` that results in the lowest cross-validation error.  
In this example, we observe diminishing values of the `CP` after reaching 5 terminal nodes (tree size $= |T|$). It's worth noting that the dashed line intersects the point where $|T| = 4$." Therefore, we can explore the potential for improved prediction accuracy by employing a pruned tree with 5 terminal nodes.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
plotcp(fit2)
```

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
prune.fit <- prune(fit2, cp[5]) 
rpart.plot(prune.fit, type = 2, digits = 3, fallen.leaves = FALSE)
```

## Step 5: Model evaluation

Recall that a smaller `Mean Squared Error (MSE)` indicates a better model.  
Therefore, we can assess the model's performance by computing the `MSE` using the test data:
```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
# predictions on the test set using the unpruned tree
yhat <- predict(fit2, newdata = test.data)
# Mean square errors
mean((yhat - test.data$medv)^2)
```

In this example, the mean square errors is 22.52. 
Note that we don't know if this value is a small or large!
We require an additional model to facilitate the comparison of `Mean Squared Error (MSE)` values across all models.

To assess whether pruning the tree can enhance the performance of this model, we compute the `Mean Squared Error (MSE)` for the `pruned tree` model using the following approach: 

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
# predictions on the test set using the pruned tree
yhat.prune <- predict(prune.fit, newdata = test.data)
# Mean square errors
mean((yhat.prune - test.data$medv)^2)
```

The square root of the MSE associated with the pruned tree is 28.13026 which is larger than that one obtained from the unpruned tree (with number of levels (or "depth") equals 6). Thus, we may conclude that the pruned tree model has lower prediction accuracy.

# Bagging and Random Forests

Here we apply bagging and random forests, using the <span style="color:blue">randomForest()</span> function in the __randomForest__ package.

## Perform bagging

The argument `mtry = 12` in the <span style="color:blue">randomForest()</span> function indicates that all 12 predictors should be considered
for each split of the tree (i.e, <span style="color:blue">bagging</span> is used).
The argument `importance = TRUE` indicates that the importance of predictors should be assessed.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
library(randomForest)
set.seed (1)
bag.boston <- randomForest(medv ~ ., 
                           data = Boston,
                           subset = Index, 
                           mtry = 12, #all 12 predictors should be considered for bagging
                           importance = TRUE)
bag.boston
```

To measure the performance of the bagged regression tree on the test set, we plot the predicted values versus the actual values. 
The plot helps us assess how well the model predicts the actual values. If the points on the plot are close to the 45-degree line (the line where predicted values equal true values), it suggests that the model is a good fit. If the points deviate significantly from this line, it indicates prediction errors.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
yhat.bag <- predict(bag.boston, newdata = test.data)
plot(yhat.bag, test.data$medv)
abline(0, 1)
mean((yhat.bag - test.data$medv)^2)
```

The test set MSE associated with the bagged regression tree is 13.03514 which is less than that one obtained from the unpruned tree (22.52).
This indicates that bagging algorithm for regression tree yielded an improvement over unpruned tree.
We could change the number of trees grown by <span style="color:blue">randomForest()</span> using the `ntree` argument:

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
bag.boston <- randomForest(medv ~ ., 
                           data = Boston,
                           subset = Index, 
                           mtry = 12, #all 12 predictors should be considered for bagging
                           ntree = 25) # number of trees are to be constructed - default = 500
yhat.bag <- predict(bag.boston, newdata = test.data)
mean((yhat.bag - test.data$medv)^2)
```

## Perform random forest

Growing a random forest proceeds in exactly the same way that we use for applying the bagged model, except that we use a smaller value of the mtry argument. 

By default, <span style="color:blue">randomForest()</span> uses $p/3$ variables when building a random forest of regression trees, and $\sqrt{p}$ variables when building a random forest of classification trees. Here we use `mtry = 6`.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
set.seed (1)
rf.boston <- randomForest(medv ~ ., 
                           data = Boston,
                           subset = Index, 
                           mtry = 6, #subset of 6 predictors - random forest algorithm
                           importance = TRUE)
yhat.rf <- predict(rf.boston, newdata = test.data)
mean((yhat.rf - test.data$medv)^2)
```

The test set MSE is 13.14; this indicates that random forests yielded an improvement over bagging in this case.

Using the <span style="color:blue">importance()</span> function, we can view the importance of each variable.

```{r,out.width="80%",fig.align="center",warning=FALSE, message=FALSE}
importance(rf.boston)
```

Two measures of variable importance are reported:

  -  The first is based upon the mean decrease of accuracy in predictions on the out of bag samples when a given variable is permuted.
  -  The second is a measure of the total decrease in node impurity that results from splits over that variable, averaged over all trees.  

In the case of regression trees, the node impurity is measured by the training `RSS`. Plots of these importance measures can be produced using the <span style="color:blue">varImpPlot()</span> function as follows:

```{r,out.width="100%",fig.align="center",warning=FALSE, message=FALSE}
varImpPlot(rf.boston, main = "Variable Importance as Measured by a Random Forest")
```
The results indicate that across all of the trees considered in the random forest, the wealth of the community (`lstat`) and the house size (`rm`) are by far the two most important variables.

